### 2014-05-31 writeups/context-dependence.tex

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{bm}
\usepackage{todonotes}

@@

@@

The situation is similar to the widely-studied problem of inference based on observation 
of a single instance of a Markov random field,
and in many cases reduces to this if we only observe the system at one time point at stationarity.
These are used, for instance, in spatial statistics \citep{besag1972nearestneighbour,gelfand2010handbook}
and image reconstruction \citep{geman1984stochastic,besag1986statistical}.
The conditioning method we consider here
is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property).
A good review of recent statistical techniques is given by \citet{friel2012bayesian}.

In the context of genomics,
it is well-known that certain short nucleotide sequences 
are in many organisms much more, or less, abundant than expected by chance \citep{karlin},
in part due to context dependence of the nucleotide mutation and substitution processes.
Phylogenetic and population genetic methods usually ignore such dependencies
in the interest of computational efficiency and lack of understanding of the underlying process,
but some progress has been made.
XXX
Usually dealt with either by using larger and larger windows (but, only up to four bp),
or by doing MCMC over additional information (entire history of changes).
Also, Markov-along-the-genome method.
Larribe and Fearnhead.

@@

The general framework also fits certain cellular automata models,
e.g.\ modeling wildfire \citep{clarke1994cellular}, 
the spread of HIV \citep{zorzenondossantos2001dynamics}, 
or land use patterns \citep{clarke1996methods}.
Complex models may introduce long-range dependencies beyond the scope of this paper,
but these methods may still prove useful in the modeling process.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'' (also known as the Gillespie algorithm \citep{vankampen}).
Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{bm}
\usepackage{todonotes}

@@

\bibliographystyle{plainnat}

@@

The situation is similar to the widely-studied problem of inference based on observation 
of a single instance of a Markov random field,
and in many cases reduces to this if we only observe the system at one time point at stationarity.
These are used, for instance, in spatial statistics \citep{besag1972nearestneighbour,gelfand2010handbook}
and image reconstruction \citep{geman1984stochastic,besag1986statistical}.
The conditioning method we consider here
is simliar to the ``coding'' scheme introduced by \citet{besag1974spatial},
that conditions on a set of sites that makes the remaining observations independent (thanks to the Gibbs property);
consistency of such methods has been shown by \citet{comets1992consistency} and reviewed by \citet{larribe2011composite}.
A good review of recent statistical techniques is given by \citet{friel2012bayesian}.

In the context of genomics,
it is well-known that certain short nucleotide sequences 
are in many organisms much more, or less, abundant than expected by chance \citep{burge1992underrepresentation},
due to the combined effects of context dependence of the nucleotide mutation process,
selective constraints on the function of the sequence,
and other processes such as biased gene conversion \citep{duret2009biased}.
Indeed, molecular studies have demonstrated that the spectrum of new mutations in humans
is highly context dependent \citep{schaibley2013influence} 
and that multinucleotide substitutions are relatively common \citep{schrider2011pervasive,terekhanova2013prevalence}.
(We are concerned with effects that are \emph{homogeneous} across the sequence,
so do not consider further the many constraints on protein-coding sequence \citep[reviewed in][]{thorne2007protein}.)
However, phylogenetic and population genetic methods usually ignore such dependencies
in the interest of computational efficiency and lack of understanding of the underlying process,
but some progress has been made.
Often, the effects of a complex mutational spectrum are explored using summary statistics of larger blocks or other sensible but \textit{ad hoc} methods.
For instance, \citet{arndt2003sequence} studied dinucleotide transitions,
and \citet{yaari2013models} displayed strong heterogeneity in the probability of synonymous mutations across 5-mers
in B-cell immunoglobin genes.
Others have made progress using simplifying assumptions \citep{berard2012accurate},
or by other approximations \citep{christensen2005pseudolikelihood}.
\citet{siepel2004phylogenetic} introduced a method to compute likelihoods under a context-dependent mutation model
by moving iteratively along the genome,
while
\citet{lunter2004nucleotide} and \citet{hobolth2008markov} used data augmentation and an MCMC algorithm to do Bayesian inference;
both these methods are promising but very computationally intensive.

@@

The general framework also fits certain cellular automata models,
e.g.\ modeling wildfire \citep{clarke1994cellular}, 
the spread of HIV \citep{zorzenondossantos2001dynamics}, 
or land use patterns \citep{wu2002calibration}.
Complex models may introduce long-range dependencies beyond the scope of this paper,
but these methods may still prove useful in the modeling process.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation''.
Briefly, this works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

\bibliography{context-dependence}
